\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}

### 2012-01-18 doc/manuscript.Rnw
Just an update.

\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

@@

\section{Abstract}
A bacterial transcriptome from the primary etiological agent of human dental
caries, \textit{Stretptococcus mutans}, is described using RNA deep sequencing
data.  We define transcripts boundaries of open reading frames based on sequence
alignments of short reads on the reference genome of UA159 from the deep
sequencing.  Based on homologous sequences and RNA secondary structures we also
report putative small RNAs, several of which are confirmed experimentally using
both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
direct role in carbon catabolite repression (CCR).  We use the developed
transcript map, open read frame, and small RNAs to determine differentially
expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
in order to refine the previously established importance of CcpA regulation as
central metabolism and virulence gene expression.  We find that more genes are
differentially expressed in the comparison between wild-type and CcpA-deletion
strains that are grown in galactose than in glucose.  We also find that more
genes differentially expressed in wild-type than in CcpA-deletion strain when
comparing glucose- and galactose-grown cells.  These two findings are different
from those from microarry experiment of \cite{Abranches2008}.  We also predicted
target genes of the small RNAs that were enriched in the following functional
categories: xxx.  We confirme that CcpA modulates the pathogenic potential of
\textit{S.\ mutans} through global control of gene expression.

@@

\section{Introduction}
Regulation of carbon catabolite repression (CCR) in \textit{Streptococcus
mutans} can involve a phosphoenolpyruvate:sugar phosphotransferase system (PTS),
histidin protein (HPr) and a transcriptional regulator, CcpA \citep{Gorke2008}.
CcpA was shown to regulate metabolism and virulence gene expression in
\textit{S.\ mutans} by \citet{Abranches2008}. They used microarray experiments
to measure expression levels of genes. Microarray experiments tend to be
qualitative. The advent of high throughput sequencing technology motivated us to
revisit the experiment to refine the result.  RNA deep sequencing data also
allowed us to define the bacterial transcriptome of \textit{S.\ mutans}. So, we
could compare putative operon expression levels as well as gene expression
levels between the wild-type (UA159) and its mutant-type (TW1) whose
\textit{ccpA} was deleted from the wild-type. This allowed us to profile
expression of genes and putative operons affected by the regulation of
\textit{ccpA}.

@@

\section{Methods and Materials}

@@

@@

>>>>

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=6,height=6}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

@@

\section{Abstract}
A bacterial transcriptome from the primary etiological agent of human dental
caries, \textit{Stretptococcus mutans}, is described using deep RNA sequencing
data.  We define transcripts boundaries of open reading frames based on sequence
alignments of short reads on the reference genome of UA159 from the deep
sequencing.  Based on homologous sequences and RNA secondary structures we also
report putative small RNAs, several of which are confirmed experimentally using
both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
direct role in carbon catabolite repression (CCR).  We use the developed
transcript map, open read frame, and small RNAs to determine differentially
expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
in order to refine the previously established importance of CcpA regulation as
central metabolism and virulence gene expression.  We find that more genes are
differentially expressed in the comparison between wild-type and CcpA-deletion
strains that are grown in galactose than in glucose.  We also find that more
genes differentially expressed in wild-type than in CcpA-deletion strain when
comparing glucose- and galactose-grown cells.  These two findings are different
from those from microarry experiment of \cite{Abranches2008}.  We also predicted
target genes of the small RNAs that were enriched in the following functional
categories: xxx.  We confirme that CcpA modulates the pathogenic potential of
\textit{S.\ mutans} through global control of gene expression.

@@

\section{Introduction}
Regulation of carbon catabolite repression (CCR) in \textit{Streptococcus
mutans} can involve a phosphoenolpyruvate:sugar phosphotransferase system (PTS),
histidin protein (HPr) and a transcriptional regulator, CcpA \citep{Gorke2008}.
CcpA was shown to regulate metabolism and virulence gene expression in
\textit{S.\ mutans} by \citet{Abranches2008}. They used microarray experiments
to measure expression levels of genes. Microarray experiments tend to be
qualitative. The advent of high throughput sequencing technology motivated us to
revisit the experiment to refine the result.  The deep RNA sequencing data also
allowed us to define the bacterial transcriptome of \textit{S.\ mutans}. So, we
could compare putative operon expression levels as well as gene expression
levels between the wild-type (UA159) and its mutant-type (TW1) whose
\textit{ccpA} was deleted from the wild-type. This allowed us to profile
expression of genes and putative operons affected by the regulation of
\textit{ccpA}.

@@

\section{Materials and Methods}

@@

Findings include:
\begin{enumerate}
\item New genes, open reading frames that are not annotated in NCBI's genome database.
\end{enumerate}

We sequenced a whole prokaryotic transcriptome of \textit{S\. mutans}, and
mapped short reads to the genomic sequence of the reference genome UA159
\citep{Ajdic2002}. 
\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
genomic sequence in order to parase \textit{B.\ anthracis} genome into a
sequence of transcripts. Most of annotated genes belong to the predicted
transcripts. We found discrepancies of annotated genes and their transcriptions:
1. A gene contains two transcripts, 
2. Non-coding regions contain transcripts.
\fixme{We could find more discrepancies. Try to find them}. 
Expression levels of adjacent genes within a transcript should be more
correlated to each other than those of adjacent genes belonging to separate transcripts
\citep{Martin2010}.

We identified a set of transcripts, albeit not a full, including small RNAs, and novel
transcripts. 
We hope that our study could serve as a foundation for a comprehensive study of
the \textit{S.\ mutans} transcriptome.
Our current knowledge of the transcriptome is limited to annotated or predicted genes.

Whole transcriptome sequencing using high throughput sequencing (HTS)
technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
transcriptome topography of \textit{S.\ mutans}.

\subsection{From Reviews}
We reconstructed the full-length transcripts by transcriptome assembly using
xxx.

We tried in vain to use short-read assemblers such as Velvet \citep{Zerbino2008} that had been developed to
assembly transcriptome.
Trinity

gene-dense transcriptomes

We ensured a high-quality transcriptome assembly.

In the data generation phase (FIG. 1a), total RNAs or mRNAs are fragmented and
converted into a library of cDNAs containing sequencing adaptors.  
In the data analysis phase (FIG. 1b), these short reads are pre-processed to
remove sequencing errors and other artefacts. The reads are subsequently
assembled to reconstruct the original RNAs and to assess their abun- dance
(‘expression counting’).

To increase the number of assem- bled transcripts, especially the less abundant
ones, ribosomal RNA (rRNA) and abundant transcripts are removed during the
first steps of library construction.
In order to retain RNAs without a poly(A) tail in the assem- bled transcriptome,
rRNA contamination can instead be removed by hybridization-based depletion
methods. \fixme{What depletion method did we use?} 
Last, the use of strand-specific RNA-seq protocols aids in the assembly and
quantification of overlapping transcripts that are derived from opposite strands
of the genome. This consideration is especially important for gene-dense
genomes, such as those of bacteria, archaea and lower eukaryotes.

Sequencing. The major factors to consider before sequencing a sample are: the
choice of sequencing plat- form, the sequencing read length and whether to use a
paired-end protocol.

Removing artefacts from RNA- seq data sets before assembly improves the read
quality, which, in turn, improves the accuracy and computa- tional efficiency of
the assembly. This step is straight- forward and can be executed using several
tools.  In general, three types of artefacts should be removed from raw RNA-seq
data: sequencing adaptors, which origi- nate from failed or short DNA
insertions during library preparation; low-complexity reads; and
near-identical reads that are derived from PCR amplification. 
Sequencing errors in NGS reads can be removed or corrected by analysing the
quality score and/or the k-mer frequency.

Reference-based transcriptome assembly is easier to perform for the simple
transcriptomes of bac- terial, archaeal and lower eukaryotic organisms, as these
organisms have few introns and little alternative splicing. Transcription
boundaries can be inferred from regions of contiguous read coverage in the
genome even with- out graph construction and traversal37,51,52. Alternative
transcription start and stop sites can also be inferred based on the 5′ cap or
poly(A) signals (if cap- or end- specific experimental protocols are used)51,53.
However, complications arise owing to the gene-dense nature of
these genomes. Many genes overlap, resulting in adja- cent genes being assembled
into one transcript, even though they are not from a polycistronic RNA. Strand-
specific RNA-seq has successfully been used to separate adjacent overlapping
genes from opposite strands in the genome51,52. Overlapping genes that are
transcribed from the same strand and that also have comparable expres- sion
levels cannot easily be separated without using cap- or end-specific RNA-seq.

In summary, reference-based assembly is generally preferable for cases in which
a high-quality reference genome already exists.

Meanwhile, experimental RNA-seq and sequencing protocols are continually
improving and should greatly reduce the informatics challenges.  RNA-seq reads
from third-generation sequencers, such as PacBio70, are longer (up to several
kilobases). PacBio sequencers are capable of sequencing a single transcript to
its full length in a single read. If this technology reaches a throughput that
is comparable to the second-generation technologies, then the need for
transcriptome assembly will probably be eliminated. Hopefully, the future of
transcriptome assembly will be ``no assembly required''.

What is the sequencing depth?

\fixme{I might have to try different filters of short reads to see how those
affect the transcript prediction.}

\subsection{My Results}

The input RNA-seq data is a list of site-wise non-negative integers that
represents a number of reads mapped to a particular genomic position
\citet{Martin2010}; a RNA-seq ``coverage'' data. We designated a gene as an
expressed if the gene belongs to a transcript with at least C2 level.

A gene

We use the term `coverage' to designate 
We verified the genes using RNA-seq data.

@@

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{manuscript-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilizing transformation of the
  count data.}
\label{figClust}
\end{figure}
<<figClust,fig=TRUE,echo=FALSE>>=
x <- rnorm(100)
hist(x)
@

### 2012-01-18 doc/manuscript.tex
Just an update.

\begin{document}

@@

\section{Prewriting about UA159 versus TW1}

\subsection{Abstract}
A bacterial transcriptome of the primary etiological agent of human dental caries,
\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
transcripts boundaries are defined. Small RNAs are predicted by computational
methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
previously established importance of CcpA regulation as central metabolism and
virulence gene expression using RNA-seq data.
CcpA is known to play a direct role in carbon catabolite repression (CCR).
We describe genes that are differentially expressed in glucose-grown wild-type
(UA159) and CcpA-deficient (TW1) strains. We found that more
genes were differentially expressed in the comparison between wild-type and
CcpA-deficient strains in galactose-grown strains than in
glucose-grown strains. How can I explain the reason?
We also found that more genes differentially expressed in wild-type than in
CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
two findings were different from those of \cite{Abranches2008}. \fixme{We may
want to see effects of different p-value cutoff.} We confirmed that CcpA
modulates the pathogenic potential of \textit{S.\ mutans} through global control
of gene expression.
We predicted a few small RNAs based on sequence alignments and RNA secondary
structures. We confirmed them experimentally. Target genes of predicted small
RNAs were predicted. The target genes were enriched in some functional
categories. We listed highly potential small RNAs with their target genes
functional categories. 
Transcripts were predicted using RNA-seq short reads mapped on the wild-type
UA159 genome.

\subsection{Reviews of papers that cited \cite{Abranches2008}}

@@

\subsubsection{Statistical analysis for differential expressions}
We used an R package, DESeq, \cite{Anders2010} to determine differentially
expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

\subsubsection{small RNAs}

RNAz is used to find small RNA candidates. Sequence alignments were a
critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
be considered to choose species related with Streptococcus mutans.
Alternatively, I find sequences similar to intergenic regions, and refine the
alignments using a multiple sequence alignment tool such as MUSCLE
\cite{Edgar2004a}. Steps for
achieving this goal include: 1. Exract intergenic regions including up- and
down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
alignments for scoring regions for small RNAs
\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
score of small RNAs. The transcripts must be defined in advance.

RNAplex
\cite{Tafer2008}

A review paper about small RNA predictions and targets
\cite{Tafer2008a}

RNAplfold
\cite{Bernhart2006}

\subsubsection{Transcripts}

``-c 10 -b 25 -force\_gp''

\subsubsection{Gene category associations and virulence genes}

We assigned genes to Gene Ontology (GO) categories by comparing the {\em
  Streptococcus} genes to bacterial proteins from the Uniref90 database
using BLASTP, and then assigning the same GO classification as the target
gene of the uniProt GOA database if the match had an $E$-value of
$<1.0\times10^{-5}.$  
A gene family was assigned a given
classification if any of its genes was assigned that classification.  To
test for associations with replacing gene transfers we used the
recombination intensity estimated for each gene, and for additive gene
transfers we used the number of additive transfers inferred for each
family.  To test for significance, we performed a Mann-Whitney $U$ test of
the values (recombination intensities or numbers of transfers) associated
with a given category vs.\ the values for the all other genes/families.  We
used the \cite{Benjamini1995} method to correct for multiple comparisons.

\subsection{Results}

@@

>>>>

\usepackage{Sweave}
\begin{document}

@@

\section{Abstract}
A bacterial transcriptome from the primary etiological agent of human dental
caries, \textit{Stretptococcus mutans}, is described using deep RNA sequencing
data.  We define transcripts boundaries of open reading frames based on sequence
alignments of short reads on the reference genome of UA159 from the deep
sequencing.  Based on homologous sequences and RNA secondary structures we also
report putative small RNAs, several of which are confirmed experimentally using
both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
direct role in carbon catabolite repression (CCR).  We use the developed
transcript map, open read frame, and small RNAs to determine differentially
expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
in order to refine the previously established importance of CcpA regulation as
central metabolism and virulence gene expression.  We find that more genes are
differentially expressed in the comparison between wild-type and CcpA-deletion
strains that are grown in galactose than in glucose.  We also find that more
genes differentially expressed in wild-type than in CcpA-deletion strain when
comparing glucose- and galactose-grown cells.  These two findings are different
from those from microarry experiment of \cite{Abranches2008}.  We also predicted
target genes of the small RNAs that were enriched in the following functional
categories: xxx.  We confirme that CcpA modulates the pathogenic potential of
\textit{S.\ mutans} through global control of gene expression.

\section{Abbreviations}

\section{Introduction}
Regulation of carbon catabolite repression (CCR) in \textit{Streptococcus
mutans} can involve a phosphoenolpyruvate:sugar phosphotransferase system (PTS),
histidin protein (HPr) and a transcriptional regulator, CcpA \citep{Gorke2008}.
CcpA was shown to regulate metabolism and virulence gene expression in
\textit{S.\ mutans} by \citet{Abranches2008}. They used microarray experiments
to measure expression levels of genes. Microarray experiments tend to be
qualitative. The advent of high throughput sequencing technology motivated us to
revisit the experiment to refine the result.  The deep RNA sequencing data also
allowed us to define the bacterial transcriptome of \textit{S.\ mutans}. So, we
could compare putative operon expression levels as well as gene expression
levels between the wild-type (UA159) and its mutant-type (TW1) whose
\textit{ccpA} was deleted from the wild-type. This allowed us to profile
expression of genes and putative operons affected by the regulation of
\textit{ccpA}.

\citet{Martin2010} developed a hidden Markov model to predict transcripts by
using site-wise expression levels. I also refined the prediction of transcripts
using multiple samples. I measured expression levels of known genes and compared
them from one experiment condition and from another. I can test whether two
genes are in the same operon or not. I assume that expression levels of two genes 
vary in similar fashion if the two genes belong to the same operon. What I mean
by similar fashion is whether they are expressed differentially with similar
fold changes. I assume that operons maps do not change between the two
experiments, or multiple experiments. This is not true. I hope that additional
samples could help to define operons. Eventually, operon maps will be determined
experimentally. Computational tools could be helpful before those ages come to
us. \fixme{This idea did not work. We should only use the prediction method.}

Although our RNA-seq data did not sampled specifically non-coding small RNAs, we
retained small RNAs in the sample by attempting to deplete only ribosomal RNAs
from the sample. Because non-coding small RNAs and coding messenger RNAs were
mixed in the sample, we sought a way of predicting small RNAs using
\program{RNAz} developed by \citet{Gruber2010}.  \program{RNAz} used homologous
sequences and RNA secondary structures to predict putative non-coding RNAs. We
confirmed the several predicted small non-coding RNAs experimentally
\citep{Altuvia2007} including the RNA-seq data set. Because small RNAs could
be associated with CCR regulation of genes (CITE), we determined differentially expressed small
RNAs between RNA samples. We found a few small RNAs to be differentially
expressed.

@@

\subsection{Statistical analysis for differential expressions}
We used an R package, DESeq, \cite{Anders2010} to determine differentially
expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

\subsection{Transcript predictions}
\citet{Martin2010} developed a hidden Markov model to predict transcripts by
using site-wise expression levels. We used the following options for the
program: ``-c 10 -b 25 -force\_gp''. We binned expression levels into 25 parts.
We allowed 10 emission states. We also used the genes as the predicted parts of
transcripts. 
Because we did not have strand-specific RNA-seq of the 13 samples due to
limited resources, we could not tell in which directions transcripts based on
RNA-seq data were oriented.

\subsection{small RNAs}
RNAz is used to find small RNA candidates. Sequence alignments were a critical
step for finding small RNAs in the approach. Phylogenies of streptococcus
species could be considered to choose species related with Streptococcus mutans.
Alternatively, I find sequences similar to intergenic regions, and refine the
alignments using a multiple sequence alignment tool such as MUSCLE
\cite{Edgar2004a}. Steps for achieving this goal include: 1. Exract intergenic
regions including up- and down-stream, 2. BLAST the intergenic regions against
non-redundant DB to find similar sequences \cite{Altschul1990}, 3. Refine each
alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the alignments for
scoring regions for small RNAs \cite{Washietl2005,Gruber2010}, and 5. Find
transcripts with high score of small RNAs. The transcripts must be defined in
advance.  We used RNAplex \citep{Tafer2008} to predict targets of putative small
RNAs. The method used RNAplfold \citep{Bernhart2006}.  Please, find a review
paper about small RNA predictions and targets by \citet{Tafer2008a}.

\subsection{Gene category associations and virulence genes}
We used two approaches for finding functional categories
\citep{Young2010,Suzuki2011}. We briefly descirbe We briefly descirbe the
prodecure of \citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO)
categories by comparing the {\em Streptococcus} genes to bacterial proteins from
the Uniref90 database using BLASTP, and then assigning the same GO
classification as the target gene of the uniProt GOA database if the match had
an $E$-value of $<1.0\times10^{-5}.$  A gene family was assigned a given
classification if any of its genes was assigned that classification.  To test
for associations with replacing gene transfers we used the recombination
intensity estimated for each gene, and for additive gene transfers we used the
number of additive transfers inferred for each family.  To test for
significance, we performed a Mann-Whitney $U$ test of the values (recombination
intensities or numbers of transfers) associated with a given category vs.\ the
values for the all other genes/families.  We used the \cite{Benjamini1995}
method to correct for multiple comparisons.

\section{Results}

Findings include:
\begin{enumerate}
\item New genes, open reading frames that are not annotated in NCBI's genome database.
\end{enumerate}

We sequenced a whole prokaryotic transcriptome of \textit{S\. mutans}, and
mapped short reads to the genomic sequence of the reference genome UA159
\citep{Ajdic2002}. 
\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
genomic sequence in order to parase \textit{B.\ anthracis} genome into a
sequence of transcripts. Most of annotated genes belong to the predicted
transcripts. We found discrepancies of annotated genes and their transcriptions:
1. A gene contains two transcripts, 
2. Non-coding regions contain transcripts.
\fixme{We could find more discrepancies. Try to find them}. 
Expression levels of adjacent genes within a transcript should be more
correlated to each other than those of adjacent genes belonging to separate transcripts
\citep{Martin2010}.

We identified a set of transcripts, albeit not a full, including small RNAs, and novel
transcripts. 
We hope that our study could serve as a foundation for a comprehensive study of
the \textit{S.\ mutans} transcriptome.
Our current knowledge of the transcriptome is limited to annotated or predicted genes.

Whole transcriptome sequencing using high throughput sequencing (HTS)
technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
transcriptome topography of \textit{S.\ mutans}.

\subsection{From Reviews}
We reconstructed the full-length transcripts by transcriptome assembly using
xxx.

We tried in vain to use short-read assemblers such as Velvet \citep{Zerbino2008} that had been developed to
assembly transcriptome.
Trinity

gene-dense transcriptomes

We ensured a high-quality transcriptome assembly.

In the data generation phase (FIG. 1a), total RNAs or mRNAs are fragmented and
converted into a library of cDNAs containing sequencing adaptors.  
In the data analysis phase (FIG. 1b), these short reads are pre-processed to
remove sequencing errors and other artefacts. The reads are subsequently
assembled to reconstruct the original RNAs and to assess their abun- dance
(‘expression counting’).

To increase the number of assem- bled transcripts, especially the less abundant
ones, ribosomal RNA (rRNA) and abundant transcripts are removed during the
first steps of library construction.
In order to retain RNAs without a poly(A) tail in the assem- bled transcriptome,
rRNA contamination can instead be removed by hybridization-based depletion
methods. \fixme{What depletion method did we use?} 
Last, the use of strand-specific RNA-seq protocols aids in the assembly and
quantification of overlapping transcripts that are derived from opposite strands
of the genome. This consideration is especially important for gene-dense
genomes, such as those of bacteria, archaea and lower eukaryotes.

Sequencing. The major factors to consider before sequencing a sample are: the
choice of sequencing plat- form, the sequencing read length and whether to use a
paired-end protocol.

Removing artefacts from RNA- seq data sets before assembly improves the read
quality, which, in turn, improves the accuracy and computa- tional efficiency of
the assembly. This step is straight- forward and can be executed using several
tools.  In general, three types of artefacts should be removed from raw RNA-seq
data: sequencing adaptors, which origi- nate from failed or short DNA
insertions during library preparation; low-complexity reads; and
near-identical reads that are derived from PCR amplification. 
Sequencing errors in NGS reads can be removed or corrected by analysing the
quality score and/or the k-mer frequency.

Reference-based transcriptome assembly is easier to perform for the simple
transcriptomes of bac- terial, archaeal and lower eukaryotic organisms, as these
organisms have few introns and little alternative splicing. Transcription
boundaries can be inferred from regions of contiguous read coverage in the
genome even with- out graph construction and traversal37,51,52. Alternative
transcription start and stop sites can also be inferred based on the 5′ cap or
poly(A) signals (if cap- or end- specific experimental protocols are used)51,53.
However, complications arise owing to the gene-dense nature of
these genomes. Many genes overlap, resulting in adja- cent genes being assembled
into one transcript, even though they are not from a polycistronic RNA. Strand-
specific RNA-seq has successfully been used to separate adjacent overlapping
genes from opposite strands in the genome51,52. Overlapping genes that are
transcribed from the same strand and that also have comparable expres- sion
levels cannot easily be separated without using cap- or end-specific RNA-seq.

In summary, reference-based assembly is generally preferable for cases in which
a high-quality reference genome already exists.

Meanwhile, experimental RNA-seq and sequencing protocols are continually
improving and should greatly reduce the informatics challenges.  RNA-seq reads
from third-generation sequencers, such as PacBio70, are longer (up to several
kilobases). PacBio sequencers are capable of sequencing a single transcript to
its full length in a single read. If this technology reaches a throughput that
is comparable to the second-generation technologies, then the need for
transcriptome assembly will probably be eliminated. Hopefully, the future of
transcriptome assembly will be ``no assembly required''.

What is the sequencing depth?

\subsection{My Results}

@@

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{manuscript-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilizing transformation of the
  count data.}
\label{figClust}
\end{figure}

### 2012-01-18 package/smutans/inst/doc/smutans.Rnw
Just an update.

% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
% \fixme{Where are flat files for genes of S. mutans UA159?}
<<samplePreparation,eval=FALSE>>=
# not run
# but create a R scirpt file with the following scripts, and
# run it, i.e., Rscript run.R
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData34( )
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
smutans.prepareDataOMZ175( ) 
smutans.prepareDataUA159CSP( ) 
smutans.prepareDataSMU86CSP( )
q("no")
@
<<options,results=hide,echo=FALSE>>=
options(digits=3)
@
Now, see some of the data objects by loading the \Rpackage{smutans}, and a data
\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
\Rpackage{DESeq} in which users should find \Rfunction{counts} and
\Rfunction{pData}. 
<<smutans,results=hide>>=
library( "smutans" )
data( "smutansGenes" )
@
There are 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA-seq samples in \Robject{smutansGenes}, which
contains the counts and also metadata about the samples:
<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
@
<<look_into_smutansGenes>>=
pData( smutansGenes )
@
The samples differ by experimental condition (``glucose'' or
``galactose'') and by library type (``ua159'' or ``tw1'').

@@

\Rpackage{GenomicRanges} might be useful in 
<<tx,eval=FALSE>>=
source("http://www.bioconductor.org/biocLite.R")
biocLite("GenomicRanges")
@

>>>>

% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
% \fixme{Where are flat files for genes of S. mutans UA159?}
<<samplePreparation,eval=FALSE>>=
# not run
# but create a R scirpt file with the following scripts, and
# run it, i.e., Rscript run.R
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData34( )
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
smutans.prepareDataOMZ175( ) 
smutans.prepareDataUA159CSP( ) 
smutans.prepareDataSMU86CSP( )
smutans.prepareTranscript( )
q("no")
@
<<options,results=hide,echo=FALSE>>=
options(digits=3)
@
Now, see some of the data objects by loading the \Rpackage{smutans}, and a data
\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
\Rpackage{DESeq} in which users should find \Rfunction{counts} and
\Rfunction{pData}. 
<<smutans,results=hide>>=
library( "smutans" )
data( "smutansGenes" )
@
There are 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA-seq samples in \Robject{smutansGenes}, which
contains the counts and also metadata about the samples:
<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
@
<<look_into_smutansGenes>>=
pData( smutansGenes )
@
The samples differ by experimental condition (``glucose'' or
``galactose'') and by library type (``ua159'' or ``tw1'').

@@

\Rpackage{GenomicRanges} is useful in summarizing predicted transcripts.
feature-genome.out-geneonly is stored in \Robject{smutans.feature.genes}.  
File FASTQ001.parsernaseq1 is ParseRNAseq output file for transcript
predictions. File FASTQ001.parsernaseq.pileup is ParseRNAseq input file for
site-wise expression levels.

\Rfunction{smutans.prepareTranscript} creates two objects:
\Robject{smutans.tx} and \Robject{smutans.txPileup}.
<<tx,eval=FALSE>>=
source("http://www.bioconductor.org/biocLite.R")
biocLite("GenomicRanges")
@
